---
title: "Bonus Lab: LinkedIn Job Postings and Cost of Living in the United States"
author: "Grace Boye-Williams and Nidhi Sathish"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## The Data
In this lab we're going to look at data regarding LinkedIn job postings from this [dataset](https://www.kaggle.com/datasets/arshkon/linkedin-job-postings) as well as a dataset containing the estimated cost of living across counties and metro areas within the United States which can be found [here](https://www.kaggle.com/datasets/asaniczka/us-cost-of-living-dataset-3171-counties). More detailed information on the datasets can be found in the [Readme](https://github.com/gracebw7/STAT331-FinalProject).

## Downloading the Data

Both datasets must be downloaded from Kaggle before running this lab. 

Steps to download:
1. Create a free account at [kaggle.com](https://www.kaggle.com) if you do not already have one.
2. Download the **LinkedIn Job Postings** dataset [here](https://www.kaggle.com/datasets/arshkon/linkedin-job-postings). From the dataset page, click the **Download** button. Unzip the downloaded folder and locate the file named `postings.csv`.
3. Download the **US Cost of Living** dataset [here](https://www.kaggle.com/datasets/asaniczka/us-cost-of-living-dataset-3171-counties). Click **Download** and locate the file named `cost_of_living_us.csv`.
4. In your RStudio project, create a folder called `data/` at the root level.
5. Move both `postings.csv` and `cost_of_living_us.csv` into the `data/` folder.


**0. Load the appropriate libraries and the data**
```{r setup}
#load libraries
library(tidyverse)
library(knitr)
library(stringr)
```

```{r load_data}
#load data
cost_of_living <- read_csv("./data/cost_of_living_us.csv")
job_postings <- read_csv("./data/postings.csv")

```

## Cleaning The Data

Before we can analyze these datasets together, both need to be cleaned. The location fields in particular are messy. `job_postings` contains a mix of formats such as `"Chicago, IL"`, `"New York, United States"`, and `"Greater Houston"`, while `cost_of_living` contains area names like `"Knoxville, TN HUD Metro FMR Area"` that need to be simplified. We will extract city and state from both datasets so that they can eventually be joined.
cost of living:

**1. Clean the jobs posting data by completing the following steps in a single pipeline. Save the result as `clean_jobs`**
- Extract the two-letter state abbreviation from the `location` column into a new column called `state`
- Extract everything before the first comma in `location` into a new column called `city`
  - Then remove everything after '-' and trim
- rename `normalized_salary` to `salary`
- Keep only the columns: `job_id`, `title`, `city`, `state`, `salary`, `currency`, `company_name`, `company_id`.
- Drop any rows where `salary`, `city`, or `state` is `NA`.

*Checkpoint: there should be 29888 observations remaining*


```{r clean jobs}
clean_jobs <- job_postings |>
  mutate(
    state = str_extract(location, "\\b[A-Z]{2}\\b"),
    city  = str_extract(location, "^[^,]+") |>
      str_remove("-.*$") |>
      str_trim()
  ) |>
  rename(salary = normalized_salary) |>
  select(job_id, title, city, state, salary, description, currency, company_name, company_id) |>
  filter(
    !is.na(salary),
    !is.na(city),
    !is.na(state)
  )
```
**2. Clean the cost of living data by completing the following steps in a single pipeline. Save the result as `clean_living`**
- Extract the two-letter state abbreviation from `areaname` into a new `state` column.
- Extract everything before the first comma in `areaname` into a new `city` column.
- Some city names are like "Los Angeles-Long Beach-Glendale", use `seperate_rows` to separate these entries into different cities based on '-'.
- Keep only rows where `family_member_count` is `"1p0c"` (a single adult with no children), then drop the `family_member_count` column.
- for each city, state there are multiple entries (from different counties within one city), summarize the costs by taking the mean for each cost.
- Keep only the columns: `state`, `city`, `housing_cost`, `food_cost`, `transportation_cost`, `healthcare_cost`, `other_necessities_cost`, `taxes`, `total_cost`, `median_family_income`.
*Checkpoint: there should be 2767 observations remaining*
``` {r clean cost}
clean_living <- cost_of_living |>
  mutate(
    state = str_extract(areaname, "\\b[A-Z]{2}\\b"),
    city  = str_extract(areaname, "^[^,]+"),
  )|>
  separate_rows(city, sep = "-") |>
  mutate(
    city = str_trim(city)
  ) |>
  filter(
    family_member_count == "1p0c"
  ) |>
   group_by(state, city) |>
   summarize(
    across(
      housing_cost:median_family_income,
      mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  select(-childcare_cost)
  
```
## Analyzing the Data

**3. Write a vector function that takes `normalized_salary` and `total_cost` as inputs and computes the ratio of the salary to the cost of living and assigned the following values based on the ratio.** Values:
- < 1: Below Cost of Living
- < 1.5: Modest
- < 2.5: Comfortable
- >= 2.5: High 

```{r salary function}
salary_tier <- function(salary, total_cost) {
  ratio <- salary / total_cost
  case_when(
    ratio < 1 ~ "Below Cost of Living",
    ratio < 1.5 ~ "Modest",
    ratio < 2.5 ~ "Comfortable",
    ratio >= 2.5 ~ "High",
  )
}
```

**4. Join the two cleaned datasets and compute a cost of living adjusted salary**
a. Join `clean_jobs` to `clean_living` using state and city
b. Add a new column `adjusted_salary` which is salary - total_cost
c. Add a new column `salary_tier` which implements the above vector function
d. Save the new dataset as `jobs_with_cost`
```{r joining data}
jobs_with_cost <- clean_jobs |>
  left_join(clean_living, by = c("state", "city")) |>
  mutate(
    adjusted_salary = (salary - total_cost),
    salary_tier = salary_tier(salary, total_cost)
  ) |>
  filter(
    !is.na(adjusted_salary)
  )
  
```