---
title: "Bonus Lab"
author: "Grace Boye-Williams and Nidhi Sathish"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## The Data
In this lab we're going to look at data regarding LinkedIn job postings from this [dataset](https://www.kaggle.com/datasets/arshkon/linkedin-job-postings) as well as a dataset containing the estimated cost of living across counties and metro areas within the United States which can be found [here](https://www.kaggle.com/datasets/asaniczka/us-cost-of-living-dataset-3171-counties). More detailed information on the datasets can be found in the [Readme]().

**0. Load the appropriate libraries and the data**
```{r setup}
#load libraries
library(tidyverse)
library(knitr)
```

```{r load_data}
#load data
cost_of_living <- read_csv("./data/cost_of_living_us.csv")
job_postings <- read_csv("./data/postings.csv")

```

##Clean The Data
cost of living:
- clean area names like "Knoxville, TN HUD Metro FMR Area" to just "Knoxville, TN"
- keep cols state, isMetro, areaname, county, family_member_count, housing_cost, food_cost, total_cost, median_family_income
- only keep rows where family_member_count is "1p0c" and then drop family member count column

job_postings:
- keep cols job_id, company_name, title, description, normalized_salary, location, company_id, currency
- some type of cleaning on location since some are city, state, country; city, state; state, country; etc.


```{r clean jobs}
clean_jobs <- job_postings |>
  mutate(
    location_raw = location,
    state = str_extract(location, "\\b[A-Z]{2}\\b"),
    city  = str_extract(location, "^[^,]+")
  ) |>
  select(job_id, title, location_raw, city, state, normalized_salary, description, currency, company_name, company_id) |>
  filter(
    !is.na(normalized_salary),
    !is.na(city),
    !is.na(state)
  )
```

``` {r clean cost}
clean_living <- cost_of_living |>
  mutate(
    state = str_extract(areaname, "\\b[A-Z]{2}\\b"),
    city  = str_extract(areaname, "^[^,]+")
  )|>
  filter(
    family_member_count == "1p0c"
  ) |>
  select(case_id, state, city, isMetro, county, housing_cost, food_cost, transportation_cost, healthcare_cost, other_necessities_cost, childcare_cost, taxes, total_cost, median_family_income)
  
```
