---
title: "Bonus Lab: LinkedIn Job Postings and Cost of Living in the United States"
author: "Grace Boye-Williams and Nidhi Sathish"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## The Data

In this lab we're going to look at data regarding LinkedIn job postings from this [dataset](https://www.kaggle.com/datasets/arshkon/linkedin-job-postings) as well as a dataset containing the estimated cost of living across counties and metro areas within the United States which can be found [here](https://www.kaggle.com/datasets/asaniczka/us-cost-of-living-dataset-3171-counties). More detailed information on the datasets can be found in the [Readme](https://github.com/gracebw7/STAT331-FinalProject).

## Downloading the Data

Both datasets must be downloaded from Kaggle before running this lab.

**Steps to download:**

1\. Create a free account at [kaggle.com](https://www.kaggle.com) if you do not already have one.

2\. Download the **LinkedIn Job Postings** dataset [here](https://www.kaggle.com/datasets/arshkon/linkedin-job-postings). From the dataset page, click the download button. Unzip the downloaded folder and locate the file named `postings.csv`. 3. Download the **US Cost of Living** dataset [here](https://www.kaggle.com/datasets/asaniczka/us-cost-of-living-dataset-3171-counties). Click download and locate the file named `cost_of_living_us.csv`.

4\. In your RStudio project, create a folder called `data/` at the root level.

5\. Move both `postings.csv` and `cost_of_living_us.csv` into the `data/` folder.

**0. Load the appropriate libraries and the data**

```{r setup}
#load libraries
library(tidyverse)
library(knitr)
library(stringr)
```

```{r}
#| label: load-data
cost_of_living <- read_csv("./data/cost_of_living_us.csv")
job_postings <- read_csv("./data/postings.csv")

```

## Cleaning The Data

Before we can analyze these datasets together, both need to be cleaned. The location fields in particular are messy. `job_postings` contains a mix of formats such as `"Chicago, IL"`, `"New York, United States"`, and `"Greater Houston"`, while `cost_of_living` contains area names like `"Knoxville, TN HUD Metro FMR Area"` that need to be simplified. We will extract city and state from both datasets so that they can eventually be joined. cost of living:

**1. To help with cleaning, create a vector function to standardize the state names**
- Create a lookup that matches state abbreviations to their full names using R's built-in `state.abb` and `state.name` constants.
- The function should accept a character vector and return the two-letter state abbreviation, handling both full state names and existing abbreviations.

:::callout-tip
Use this helper function inside the cleaning pipelines below, it uses R's built in state constants.
:::
```{r}
#| label: helper-function
extract_state <- function(text_chunk) {
  # Create the dictionary using built-in R constants
  state_lookup <- state.abb
  names(state_lookup) <- state.name
    case_when(
      # If it's a full state name, look up its abbreviation
      text_chunk %in% names(state_lookup) ~ state_lookup[text_chunk],
      
      # If it's already a 2-letter uppercase word, extract it
      str_detect(text_chunk, "\\b[A-Z]{2}\\b") ~ str_extract(text_chunk, "\\b[A-Z]{2}\\b"),
      
      # Default if not a state entry
      TRUE ~ NA_character_
    )
  }
```

**2. Clean the jobs posting data by completing the following steps in a single pipeline.** Save the result as **`clean_jobs`**

\- Extract 2 strings from the `location` column, use the helper above to find which part contains the state, if the state is the second part, the first part must be the city, if the state is the first part, the city is `NA`
\- Extract the city and state information into 2 new columns `state` and `city`
\- rename `normalized_salary` to `salary`
\- Keep only the columns: `job_id`, `title`, `city`, `state`, `salary`, `currency`, `company_name`, `company_id`.
\- Drop any rows where `salary`, `city`, or `state` is `NA`.

*Checkpoint: there should be 29888 observations remaining*

```{r}
#| label: clean-jobs
clean_jobs <- job_postings |>
  
  mutate(
    # Isolate parts to verify if the first part was actually a state name
    part_before = str_extract(location, "^[^,]+") |> str_trim(),
  
    part_after = str_extract(location, "[^,]+$") |> str_trim(),
    
    state = coalesce(
      extract_state(part_after),
      extract_state(part_before)
    ),
    
    # If part_before matches a state name or code, then no city exists.
    # Otherwise, clean the city name.
    city = case_when(
      !is.na(extract_state(part_before)) ~ NA_character_,
      TRUE ~ part_before |> str_remove("-.*$") |> str_trim()
    )
  )|>
  rename(salary = normalized_salary) |>
  select(job_id, title, city, state, salary, currency, company_name, company_id) |>
  filter(
    !is.na(salary),
    !is.na(city),
    !is.na(state)
  )
```


**3. Clean the cost of living data by completing the following steps in a single pipeline.** Save the result as **`clean_living`**

\- Extract the two-letter state abbreviation from `areaname` into a new `state` column.

\- Extract everything before the first comma in `areaname` into a new `city` column.

\- Some city names are like "Los Angeles-Long Beach-Glendale", use `seperate_rows` to separate these entries into different cities based on '-' [hint](https://tidyr.tidyverse.org/reference/separate_rows.html)

\- Keep only rows where `family_member_count` is `"1p0c"` (a single adult with no children), then drop the `family_member_count` column.

\- for each city, state there are multiple entries (from different counties within one city), summarize the costs by taking the mean for each cost.

\- Keep only the columns: `state`, `city`, `housing_cost`, `food_cost`, `transportation_cost`, `healthcare_cost`, `other_necessities_cost`, `taxes`, `total_cost`, `median_family_income`.

*Checkpoint: there should be 2,767 observations remaining*

```{r}
#| label: clean-living
clean_living <- cost_of_living |>
  mutate(
    state = str_extract(areaname, "\\b[A-Z]{2}\\b"),
    city  = str_extract(areaname, "^[^,]+"),
  )|>
  separate_rows(city, sep = "-") |>
  mutate(
    city = str_trim(city)
  ) |>
  filter(
    family_member_count == "1p0c"
  ) |>
   group_by(state, city) |>
   summarize(
    across(
      housing_cost:median_family_income,
      mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  select(-childcare_cost)
  
```

## Analyzing the Data

**4. Write a vector function that takes `normalized_salary` and `total_cost` as inputs and computes the ratio of the salary to the cost of living and assigned the following values based on the ratio.**

Values:

-   \< 1: Below Cost of Living

-   \< 1.5: Modest

-   \< 2.5: Comfortable

-   \>= 2.5: High

```{r}
#| label: salary-tier-function
salary_tier <- function(salary, total_cost) {
  ratio <- salary / total_cost
  case_when(
    ratio < 1 ~ "Below Cost of Living",
    ratio < 1.5 ~ "Modest",
    ratio < 2.5 ~ "Comfortable",
    ratio >= 2.5 ~ "High",
  )
}
```

**4. Join the two cleaned datasets and compute a cost of living adjusted salary**

a\. Join `clean_jobs` to `clean_living` using state and city

b\. Add a new column `adjusted_salary` which is salary - total_cost

c\. Add a new column `salary_tier` which implements the above vector function d. Save the new dataset as `jobs_with_cost`

```{r}
#| label: join-data
jobs_with_cost <- clean_jobs |>
  left_join(clean_living, by = c("state", "city")) |>
  mutate(
    adjusted_salary = (salary - total_cost),
    salary_tier = salary_tier(salary, total_cost)
  ) |>
  filter(
    !is.na(adjusted_salary)
  )
```
